Spike spectra for recurrences

### 2022-08-04 Manuscript/inter_spike_manuscript_mdpi.tex
add orcid
minor issues

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0002-9943-5391} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{K. Hauke Kraemer $^{1,}$*\orcidA{}, Frank Hellmann $^{1,}$, Mehrnaz Anvari $^{1}$, J\"urgen Kurths $^{1,2,4}$ and Norbert Marwan $^{1,3,}$*}

@@

The signal, which we would like to transform is decomposed into a set of appropriate basis functions.\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} Instead of using trigonometric 
functions, as it is the idea in the Fourier decomposition, we use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} \bfX_{i,j} &= \sum_{k=0}^{N}\delta\left(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \right), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n \quad,~s.t.~~ n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}
\todo[inline]{HK: Eq.\eqref{eq_basis_matrix} is somehow ugly. Maybe someone can get rid of that sum and make it nicer? I tried with 
modulo, but just could not figure it out exactly. There must be an easy way I do not see.}
Note that due to the shifting of each of the basis functions of inter-spike period $T_\text{is}$, $\bfX$ is not linear independent anymore. 
Furthermore, there will be identical basis functions and also basis functions, which do not allow for an unambiguous inter-spike period, 
if we would include all $N$ possible inter-spike periods for a signal of length $N$ instead of $\ceil*{N/2}+1$ (Fig.~\ref{fig_tau_rr_basis_functions}A). 
The reason is that in contrast to a trigonometric decomposition, where the Nyquist frequency marks a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$ (schematically illustrated in 
Fig.~\ref{fig_tau_rr_basis_functions}B).\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Inter spike spectra of power grid frequency data}\label{sec_power_grid}

@@

\noindent Then we construct the inter-spike spectrogram of the $\tau$-RR by first determining an appropriate embedding. By using a recent Tree-Embedding-Ansatz \cite{Kraemer2022} we minimize the 
\textit{false nearest neighbor} statistic \cite{hegger1999} and use the continuity statistic \cite{pecora2007} for potential delays. Eventually we obtain an $8$-dimensional embedding 
with delays $\tau s=[0, 18, 36, 48, 62, 101, 114]$ (in sampling units) for the entire time series. Similar to the preceding approach we embed the time series with these embedding parameters 
on windows of size $ws=1,200$ and a unity shift.\footnote{We use a slightly larger window here than in the short FT, because the embedding causes a ``loss'' of data points and we want to 
cover similar time spans.} The RPs are computed on these embedded trajectories with a fixed recurrence threshold corresponding to $10\%$ global recurrence rate \cite{kraemer2018} 
and the inter-spike spectra of the corresponding $\tau$-RRs is obtained by using STLS regression (see Section \ref{sec_tau_rr_method}) and a regularization threshold corresponding to 
$\rho=0.9$ accordance of $\tau$-RRs and re-composed signals. We only use the first $200$ data points of the $\tau$-RR (covering a time span of $1$ Myr). 
The spectrogram also highlights the $95$, $124$ and $405$ kyr periods as expected (Fig.~\ref{fig_laskar_spectra}B). 
Additional power is distributed in its harmonics at $190$ and $248$ kyr periods. Finally, the standard FT with a Hamming window of the same $\tau$-RRs used to obtain the inter-spike 
spectrogram in panel B, yields a smeared spectrogram which does not reflect the expected periods, but rather suffers from the spike train behavior, i.e. many excited harmonics, of the 
FT described in Section \ref{sec_tau_rr_intro} and Fig.~\ref{fig_tau_rr_dirac_comb}. The shown results are robust to a change of embedding parameters and windowsizes. However, a too low 
regularization threshold smears the clear spectrogram in panel B.

@@

We have successfully used the idea of transforming the $\tau$-RR for the detection of bifurcations in the Logistic map. By constructing appropriate surrogates of the inter spike spectra, 
and thus a null model, the number of significant peaks in the inter spike spectrum correlated well with the positive Lyapunov exponent. This measure was also able to resolve 
period-doubling bifurcations. However, we have to admit this way the detection of a bifurcation is only possible when the additional period(s) is not an integer multiple of the former period(s). 
This behavior is described in the application to the R\"ossler system, where we explicitly showed the different inter spike power spectra for period-2, period-3 and chaotic dynamics.
Further development might potentially incorporate the mutual distance of peaks in the spectrum for a better correlation to the Lyapunov exponent. The inter spike spectra of power grid 
frequency data illustrate that our proposed method may serve as a valuable source of information in addition to a standard Fourier analysis. And last but not least, we showed that our approach 
really incorporates frequencies, which are apparent in the Fourier spectrum of the signal, by applying it to analytically derived eccentricity data, where the dominant frequencies are well 
known.

We discuss some more technical details in the following, which will also affect any application of the proposed method. 
First of all, the number of required basis functions 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ for an input signal of length $N$ is the crucial bottleneck of this approach, which is why it does not show good scaling behavior. The subsequent sparse 
regression, therefore, gets computationally intensive for $N>1,000$. Depending on the memory of the computer used, input signals $N>2,000$ usually do not work anymore. This means that signals 
often need to get downsampled as a preprocessing step (e.g., see Appendix \ref{sec_power_grid_appendix}). Second, the regularization parameter $\alpha$ for the regression is a crucial free parameter. 
As described in Section \ref{sec_tau_rr_method} our idea in this paper is to select the $\alpha$ such that the re-composed signal $\tilde{\bf{s}}=\bf{X}^T\hat{\boldsymbol\beta}$ matches a 
given (Pearson) correlation coefficient $\rho_{\bf{s},\tilde{\bf{s}}}$ between the original signal $\bf{s}$ and itself. This ensures that $\alpha$ adjusts itself to the data as well as to the 
used regression method. We found that this increases the comparability of different spectra, especially when performing a running window approach in order to obtain an evolutionary spectrogram 
(Fig.~\ref{fig_laskar_spectra}). 
However, the two different sparse regression algorithms we encountered in this article (LASSO and STLS) yield different results for the same desired $\rho_{\bf{s},\tilde{\bf{s}}}$. 
Even if the spectra obtained in this way look qualitatively similar, they are not always quantitatively similar. The reason for this 
is that $\rho_{\bf{s},\tilde{\bf{s}}}$ is not a smooth function of $\alpha$ in case of STLS, due to the hard-thresholding involved \cite{Brunton2016}, which is shown in Figure~\ref{fig_regularization}.
Third, when adopting our idea of applying the inter spike spectrum to the $\tau$-recurrence rate of the signals state space trajectory, the embedding process induces additional free parameters. 
This is not a drawback of the proposed method, but rather a drawback of the particular application method that we have heavily used in this article and which was the main motivation for developing 
the proposed method. As a very last remark, we draw attention to the fact that sparse regression can be transformed into sparse logistic regression when the signal we would like to transform is binary.

@@

A novel type of powerspectrum, the \textit{inter-spike spectrum}, has been proposed. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution (we chose LASSO or a sequentially 
thresholded least squares regression STLS). Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the period (or frequency) the inter spike spectrum is obtained. 
Although this novel power spectrum is superior to an ordinary FFT-based power spectrum when the signal has a spike-train-like appearance, the authors suggest that this method be considered 
as an additional source of information and not as a substitute for ordinary Fourier analysis. 
Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.

The invention of the proposed method has been motivated by the idea of transforming $\tau$-recurrence rate signals ($\tau$-RRs) into their frequency domain. 
This general idea \cite{Zbilut2008} allows for a frequency analysis of high dimensional systems, because the RP is a representation of the system's state space trajectory.   
The $\tau$-RR of a recurrence plot (RP) usually has a spiky shape, especially for map-like data, and the inter-spike spectrum can reliably reveal the system's dominant frequencies, 
which is not possible when Fourier transforming the $\tau$-RR or the underlying signal itself. Since the position of the peaks 
in the $\tau$-RR are not sensitive to noise, the corresponding inter spike spectrum also yields robust results in the presence of noise.

@@

\section{Inter spike spectra for noisy R\"ossler system}

\begin{figure}[h!]
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_tau_rr_example_roessler_noise}
 \caption{Same as in Fig.~\ref{fig_tau_rr_example_roessler}, but here with 5\% additive Gaussian white noise on each component $x$, $y$ and $z$. The appearance of the inter spike spectra 
 in \textbf{J, K, L}, and the Fourier spectra in \textbf{M, N, P} are unaffected by the additive noise.}
\label{fig_tau_rr_example_roessler_noise}
\end{figure}

@@

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_regularization}
 \caption{Pearson correlation coefficient $\rho_{\bf{s},\tilde{\bf{s}}}$ between the input signal signal $\bf{s}$ and the re-composed signal $\tilde{\bf{s}}=\bf{X}^T\hat{\boldsymbol\beta}$ as a 
 function of the regularization parameter $\alpha$ for \textbf{A} LASSO and \textbf{B} sequentially thresholded least squares (STLS) regression (see also Section \ref{sec_tau_rr_method}). 
 The input time series is of length $N=200$ and stems from the $\tau$-RR of R\"ossler system in regular dynamics.}
\label{fig_regularization}
\end{figure}

>>>>

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0002-9943-5391} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorB}{0000-0001-6643-9508} % Add \orcidB{} behind the author's name
\newcommand{\orcidauthorC}{0000-0003-1437-7039} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{K. Hauke Kraemer $^{1,}$*\orcidA{}, Frank Hellmann $^{1,}$, Mehrnaz Anvari $^{1}$\orcidB{}, J\"urgen Kurths $^{1,2,4}$ and Norbert Marwan $^{1,2,3,}$*\orcidC{}}

@@

The signal, which we would like to transform is decomposed into a set of appropriate basis functions.\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} Instead of using trigonometric 
functions, as it is the idea in the Fourier decomposition, we use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad s.t.~~ n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}
\todo[inline]{HK: Eq.\eqref{eq_basis_matrix} is somehow ugly. Maybe someone can get rid of that sum and make it nicer? I tried with 
modulo, but just could not figure it out exactly. There must be an easy way I do not see.}
Note that due to the shifting of each of the basis functions of inter-spike period $T_\text{is}$, $\bfX$ is not linear independent anymore. 
Furthermore, there will be identical basis functions and also basis functions, which do not allow for an unambiguous inter-spike period, 
if we would include all $N$ possible inter-spike periods for a signal of length $N$ instead of $\ceil*{N/2}+1$ (Fig.~\ref{fig_tau_rr_basis_functions}A). 
The reason is that in contrast to a trigonometric decomposition, where the Nyquist frequency marks a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$ (schematically illustrated in 
Fig.~\ref{fig_tau_rr_basis_functions}B).\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{inter-spike spectra of power grid frequency data}\label{sec_power_grid}

@@

\noindent Then we construct the inter-spike spectrogram of the $\tau$-RR by first determining an appropriate embedding. By using a recent Tree-Embedding-Ansatz \cite{Kraemer2022} we minimize the 
\textit{false nearest neighbor} statistic \cite{hegger1999} and use the continuity statistic \cite{pecora2007} for potential delays. Eventually we obtain an $8$-dimensional embedding 
with delays $\tau s=[0, 18, 36, 48, 62, 101, 114]$ (in sampling units) for the entire time series. Similar to the preceding approach we embed the time series with these embedding parameters 
on windows of size $ws=1,200$ and a unity shift.\footnote{We use a slightly larger window here than in the short FT, because the embedding causes a ``loss'' of data points and we want to 
cover similar time spans.} The RPs are computed on these embedded trajectories with a fixed recurrence threshold corresponding to $10\%$ global recurrence rate \cite{kraemer2018} 
and the inter-spike spectra of the corresponding $\tau$-RRs is obtained by using STLS regression (see Section \ref{sec_tau_rr_method}) and a regularization threshold corresponding to 
$\rho=0.9$ accordance of $\tau$-RRs and re-composed signals. We only use the first $200$ data points of the $\tau$-RR (covering a time span of $1$ Myr). 
The spectrogram also highlights the $95$, $124$ and $405$ kyr periods as expected (Fig.~\ref{fig_laskar_spectra}B). 
Additional power is distributed in its harmonics at $190$ and $248$ kyr periods. Finally, the standard FT with a Hamming window of the same $\tau$-RRs used to obtain the inter-spike 
spectrogram in panel B, yields a smeared spectrogram which does not reflect the expected periods, but rather suffers from the spike train behavior, i.e., many excited harmonics, of the 
FT described in Section \ref{sec_tau_rr_intro} and Fig.~\ref{fig_tau_rr_dirac_comb}. The shown results are robust to a change of embedding parameters and windowsizes. However, a too low 
regularization threshold smears the clear spectrogram in panel B.

@@

We have successfully used the idea of transforming the $\tau$-RR for the detection of bifurcations in the Logistic map. By constructing appropriate surrogates of the inter-spike spectra, 
and thus a null model, the number of significant peaks in the inter-spike spectrum correlated well with the positive Lyapunov exponent. This measure was also able to resolve 
period-doubling bifurcations. However, we have to admit this way the detection of a bifurcation is only possible when the additional period(s) is not an integer multiple of the former period(s). 
This behavior is described in the application to the R\"ossler system, where we explicitly showed the different inter-spike power spectra for period-2, period-3, and chaotic dynamics.
Further development might potentially incorporate the mutual distance of peaks in the spectrum for a better correlation to the Lyapunov exponent. The inter-spike spectra of power grid 
frequency data illustrate that our proposed method may serve as a valuable source of information in addition to a standard Fourier analysis. And last but not least, we showed that our approach 
really incorporates frequencies, which are apparent in the Fourier spectrum of the signal, by applying it to analytically derived eccentricity data, where the dominant frequencies are well 
known.

We discuss some more technical details in the following, which will also affect any application of the proposed method. 
First of all, the number of required basis functions 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ for an input signal of length $N$ is the crucial bottleneck of this approach, which is why it does not show good scaling behavior. The subsequent sparse 
regression, therefore, gets computationally intensive for $N>1,000$. Depending on the memory of the computer used, input signals $N>2,000$ usually do not work anymore. This means that signals 
often need to get downsampled as a preprocessing step (e.g., see Appendix \ref{sec_power_grid_appendix}). Second, the regularization parameter $\alpha$ for the regression is a crucial free parameter. 
As described in Sect.~\ref{sec_tau_rr_method}, our idea in this paper is to select the $\alpha$ such that the re-composed signal $\tilde{\vec{s}}=\bfX^T\hat{\boldsymbol\beta}$ matches a 
given (Pearson) correlation coefficient $\rho_{\vec{s},\tilde{\vec{s}}}$ between the original signal $\vec{s}$ and itself. This ensures that $\alpha$ adjusts itself to the data as well as to the 
used regression method. We found that this increases the comparability of different spectra, especially when performing a running window approach in order to obtain an evolutionary spectrogram 
(Fig.~\ref{fig_laskar_spectra}). 
However, the two different sparse regression algorithms we encountered in this article (LASSO and STLS) yield different results for the same desired $\rho_{\vec{s},\tilde{\vec{s}}}$. 
Even if the spectra obtained in this way look qualitatively similar, they are not always quantitatively similar. The reason for this 
is that $\rho_{\vec{s},\tilde{\vec{s}}}$ is not a smooth function of $\alpha$ in case of STLS, due to the hard-thresholding involved \cite{Brunton2016}, which is shown in Fig.~\ref{fig_regularization}.
Third, when adopting our idea of applying the inter-spike spectrum to the $\tau$-recurrence rate of the signals state space trajectory, the embedding process induces additional free parameters. 
This is not a drawback of the proposed method, but rather a drawback of the particular application method that we have heavily used in this article and which was the main motivation for developing 
the proposed method. As a very last remark, we draw attention to the fact that sparse regression can be transformed into sparse logistic regression when the signal we would like to transform is binary.

@@

A novel type of powerspectrum, the \textit{inter-spike spectrum}, has been proposed. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution (we chose LASSO or a sequentially 
thresholded least squares regression STLS). Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the period (or frequency) the inter-spike spectrum is obtained. 
Although this novel power spectrum is superior to an ordinary FFT-based power spectrum when the signal has a spike-train-like appearance, the authors suggest that this method be considered 
as an additional source of information and not as a substitute for ordinary Fourier analysis. 
Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter-spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.

The invention of the proposed method has been motivated by the idea of transforming $\tau$-recurrence rate signals ($\tau$-RRs) into their frequency domain. 
This general idea \cite{Zbilut2008} allows for a frequency analysis of high dimensional systems, because the RP is a representation of the system's state space trajectory.   
The $\tau$-RR of a recurrence plot (RP) usually has a spiky shape, especially for map-like data, and the inter-spike spectrum can reliably reveal the system's dominant frequencies, 
which is not possible when Fourier transforming the $\tau$-RR or the underlying signal itself. Since the position of the peaks 
in the $\tau$-RR are not sensitive to noise, the corresponding inter-spike spectrum also yields robust results in the presence of noise.

@@

\section{inter-spike spectra for noisy R\"ossler system}

\begin{figure}[h!]
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_tau_rr_example_roessler_noise}
 \caption{Same as in Fig.~\ref{fig_tau_rr_example_roessler}, but here with 5\% additive Gaussian white noise on each component $x$, $y$ and $z$. The appearance of the inter-spike spectra 
 in \textbf{J, K, L}, and the Fourier spectra in \textbf{M, N, P} are unaffected by the additive noise.}
\label{fig_tau_rr_example_roessler_noise}
\end{figure}

@@

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_regularization}
 \caption{Pearson correlation coefficient $\rho_{\vec{s},\tilde{\vec{s}}}$ between the input signal signal $\vec{s}$ and the re-composed signal $\tilde{\vec{s}}=\bfX^T\hat{\boldsymbol\beta}$ as a 
 function of the regularization parameter $\alpha$ for \textbf{A} LASSO and \textbf{B} sequentially thresholded least squares (STLS) regression (see also Section \ref{sec_tau_rr_method}). 
 The input time series is of length $N=200$ and stems from the $\tau$-RR of R\"ossler system in regular dynamics.}
\label{fig_regularization}
\end{figure}
